---
title: "Meta-model Results"
author: "Xijin"
date: "10/22/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Motivation

Based on one of those simulation settings and simulate 1000 times to see the results and the result from meta-model

# Simulation settings:

1. Modelling strategies: maximum likelihood

2. Distribution of predictor variables: mulivariate normal distribution

3. Sample size: 400

4. Number of candidate predictors P: 4

5. Events fraction: 0.5

6. Predictor effects: equal effects


functions 

```{r,warning=FALSE,error=FALSE,include=FALSE}
require(MASS) # MVN function
require(rms)  #lrm function and predict.lrm function
```


```{r,warning=FALSE,error=FALSE}
#predictive error estimation by meta model
mspe_meta <- function(n,ef,P){
  exp( -0.59 - 1.06*log(n) + 0.36*log(ef) + 0.94*log(P))
}

brier_meta <- function(n,ef,P){
  exp(-0.91 - 0.04*log(n) + 0.62*log(ef) + 0.04*log(P) )
}

mape_meta <- function(n,ef,P){
  exp( -0.48 - 0.53*log(n) + 0.31*log(ef) + 0.48*log(P))
}

#predictive error real values
brier_true <- function(p,SIMxy){
  sum((SIMxy[,"y"]-p)^2)/length(p)
}

mspe_true <- function(p,SIMxy,dgm.par){
  design.mat <- as.matrix(cbind(1,SIMxy[,-which(colnames(SIMxy)=="y")]))
  p_true <- exp(design.mat%*%dgm.par)/(1+exp(design.mat%*%dgm.par))
  mean((p_true-p)^2)
}

mape_true <- function(p,SIMxy,dgm.par){
  design.mat <- as.matrix(cbind(1,SIMxy[,-which(colnames(SIMxy)=="y")]))
  p_true <- exp(design.mat%*%dgm.par)/(1+exp(design.mat%*%dgm.par))
  mean(abs(p_true-p))
}

#function for prediction
  
  predict.lrm <- function(lp){  
    as.vector(exp(lp)/(1+exp(lp)))
  } 

#generate independent variables
gen.MVNX <- function(n,mu,sigma){
  mvrnorm(n=n,mu=mu,Sigma=sigma)
}

#generate dependent variables
gen.binY <- function(SIMx,dgm.par){
  no.X <- ncol(SIMx)
  design.mat <- cbind(1,SIMx)
  p <- exp(design.mat%*%dgm.par)/(1+exp(design.mat%*%dgm.par))
  y <- rbinom(length(p),1,p)
  SIMxy <-data.frame(x=SIMx,y=y)
  colnames(SIMxy)[1:no.X]<-paste("X",1:no.X,sep="")  
  SIMxy
}
```

```{r}
#data generation
args <- list(
  mu = c(0,0,0,0),
  sigma = matrix(c(1,0,0,0,
                    0,1,0,0,
                    0,0,1,0,
                    0,0,0,1),nrow=4),
  n = 400
)
datagen <- "gen.MVNX"
dgm.par <- c(-0.0009684,  0.2832211,  0.2832211,  0.2832211,  0.2832211)
```



# Results based on 1000 simulations

```{r}
sim.num <- 1000 # 1000 simulations
brier_true_values <- rep(0,sim.num)
mape_true_values <- rep(0,sim.num)
mspe_true_values <- rep(0,sim.num)
ef <- rep(0,sim.num)

for (i in 1:sim.num){
  #data generation, this is original data set 
  SIMx <- do.call(what=datagen,args=args)
  SIMxy <- gen.binY(SIMx=SIMx,dgm.par=dgm.par)
  SIMx <- SIMxy[,-ncol(SIMxy)]
  y <- SIMxy[,"y"]
  ef[i] <- sum(y)/length(y)
  #model fitting and corresponding predicted values
  formu <- as.formula(paste("y~",
                            paste(colnames(SIMxy)[-which(colnames(SIMxy)=="y")],collapse="+"),sep=""))
  fit <- lrm(formu,data=SIMxy)
  #predict outcomes
  SIMx <- cbind("Intercept"=1,SIMx)
  lp <- unlist(as.matrix(SIMx[,names(coef(fit))])%*%data.matrix(coef(fit)))
  pred <- predict.lrm(lp)
  #metrics calculation
  brier_true_values[i] <- brier_true(p=pred,SIMxy)
  mape_true_values[i] <- mape_true(p=pred,SIMxy,dgm.par)
  mspe_true_values[i] <- mspe_true(p=pred,SIMxy,dgm.par)
}
```


# Results based on meta-model
```{r}
brier_meta_values <- brier_meta(n=args$n,ef=0.5,P=length(dgm.par))
mape_meta_values <- mape_meta(n=args$n,ef=0.5,P=length(dgm.par))
mspe_meta_values <- mspe_meta(n=args$n,ef=0.5,P=length(dgm.par))
```


# Results comparison
```{r}
dat_sim <- data.frame(cbind(Brier=brier_true_values,
                            MAPE=mape_true_values,
                            MSPE=mspe_true_values))
library(reshape2)
dat_sim <- melt(dat_sim,
              measure.vars=c('Brier',
                             'MAPE',
                             'MSPE'))
dat_sim$variable <- as.factor(dat_sim$variable)
meta_values <- c(brier_meta_values,mape_meta_values,mspe_meta_values)
dat <- cbind(dat_sim,meta=c(rep(brier_meta_values,sim.num),
                            rep(mape_meta_values,sim.num),
                            rep(mspe_meta_values,sim.num)))
ggplot(dat) +
  geom_boxplot(aes(x=variable, y=value))+
  geom_point(aes(x=variable, y=value))+
  geom_hline(aes(yintercept=meta_values[1],col="Metamodel"))+
  geom_hline(aes(yintercept=meta_values[2],col="Metamodel"))+
  geom_hline(aes(yintercept=meta_values[3],col="Metamodel"))+
  theme_bw()+
 # labs(color = "Meta model result",
#       x="Prediction error metrics",
#       ylab="Value")+
  theme(legend.text = element_text(NULL))+
  scale_colour_manual(NULL,values=c('Metamodel' = "red"))

dat_diff <- dat_sim
meta <- c(rep(brier_meta_values,sim.num),
                            rep(mape_meta_values,sim.num),
                            rep(mspe_meta_values,sim.num))
dat_diff$value <- dat_sim$value-meta
ggplot(dat_diff) +
  geom_boxplot(aes(x=variable, y=value))+
  geom_point(aes(x=variable, y=value),alpha = 0.03)+
  geom_hline(aes(yintercept=0,col="Metamodel"))+
  scale_colour_manual(NULL,values=c('Metamodel' = "red"))+
   theme_bw()+
  theme(legend.text = element_text(NULL))

```

